A fast GPU solver for near-Toeplitz tridiagonal systems

### 2015-11-04 paper-draft/paper.tex
update paper

\usepackage{amsmath,bm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{minted}
\usepackage{xcolor}
\usepackage{multirow}

@@

\begin{frontmatter}
\author{Ashwin Srinath}
\title{A fast GPU solver for near-Toeplitz tridiagonal systems}
\maketitle

\begin{abstract}
    In this paper, we consider the solution of linear
    systems of the form $A\bm{x} = \bm{d}$,
    where $A$ is \emph{near-Toeplitz}.
    These systems appear in numerical schemes
    such as
    Alternating Direction Implicit (ADI) methods,
    compact finite differences,
    and one-dimensional ordinary
    and partial differential equations.
    We develop a GPU solver that takes advantage
    of the known structure of the matrix
    to reduce the number of computations
    and improve memory access.
    We also describe the implementation of a multi-GPU solver
    using MPI and CUDA.
    We describe two applications:
    2-D ADI applied to the heat equation on a single GPU,
    and computation of
    3-D compact finite differences on multiple GPUs.
    Our solver is able to achieve 6x speedup
    over an Intel MKL solver running on a single core,
    and 2x speedup
    over the NVIDIA CUSPARSE solver.
    Additionally, we demonstrate scalability up to 64 GPUs.

@@

This paper is organized as follows:
Section \ref{sec:preliminaries}
describes algorithms for solving tridiagonal systems on GPUs.
The cyclic reduction algorithm in particular, is discussed in detail,
and issues with its implementation on GPUs is discussed.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=300pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Storing forward reduction coefficients
    from each reduction step $m$}
\label{fig:cyclic-reduction-precomputing}
\end{figure}

@@

In this section, we describe the implementation of a GPU solver
for solving a given near-Toeplitz tridiagonal system
for $N_{rhs}$ right hand sides.

We describe two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
In both approaches,
we map each of the $N_{rhs}$ tridiagonal systems
to blocks,
and equations or \emph{indices} of a system
to individual threads within a block.

Following the algorithm described in \ref{sec:proposed-algorithm},
the forward reduction coefficients
are first computed and loaded into global memory.
The right hand sides are stored in a single
contiguous region global memory.

@@

In the global memory approach,
we launch several kernels---one
for each single forward reduction step,
an each backward substitution step.
The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory directly.
Because the coefficients
$k_{1,i}$ and $k_{2,i}$
required in the forward reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
are precomputed and available in global memory,
we do not compute
$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
$k_{1,i}$ and $k_{2,i}$.
This reduces the number of uncoalesced accesses to global memory.
In the backward substitution steps
(Equation \ref{eqn:precomputed-backward-substitution-step}),
the coefficients $a^m$, $b^m$ and $c^m$
are accessed from global memory without coalescing problems.

The advantage of the global memory approach is that
very large tridiagonal systems can be solved without
being limited by the available shared memory per block.
Further,
because we launch a kernel for each
forward reduction and back substitution step,
we can specify the number of threads per block at each launch so that
there are no idle threads throughout the solution process.
($n/2$ threads per block for the first forward reduction step,
$n/4$ for the second, and so on).
The number of blocks launched is equal to the number of right
hand sides.
Exit from a kernel marks a global synchronization point,
so no explicit synchronization is required within the kernel.

@@

In the shared memory approach,
we launch a single kernel with $n/2$ threads per block,
and as many blocks as right hand sides.
Each block allocates shared memory with size $n/2$.
We note that the arrays containing the precomputed coefficients,
are kept in global memory.
Each thread performs the first reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding uncoalesced global memory accesses.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completeley
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
Explicit synchronization is required at the end of each step.

\pagebreak
%----------------------------------------------------------------------%
\section{Results} \label{sec:results-single-gpu}
%----------------------------------------------------------------------%

In this section we present a performance overview
of our Near-Toeplitz tridiagonal solver (NEATO)
against a multi-threaded Intel MKL solver and
the CUSPARSE GPU solver
(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).

The CPU code is compiled with the Intel C++ compiler
and run on up to 16 cores of the
Intel Xeon Processor E5-2670 v2 (25M Cache, 2.50 GHz).
GPU code is compiled with the CUDA toolkit (version 6.5.14),
and run on the
NVIDIA Tesla K20 Accelerator.
\texttt{-O2} level optimizations are turned on for both
CPU and GPU code.

@@

\begin{figure}[!htb]
    \centering
    \begin{minipage}{0.75\textwidth}
        \centering
        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
        \caption{Comparison of global memory and shared
            memory implementations of NEATO (2D problems)}
        \label{fig:global-vs-shared-2d}
    \end{minipage}
    \\
    \begin{minipage}{0.75\textwidth}
        \centering
        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
        \caption{Comparison of global memory and shared memory
            implementations of NEATO (2D problems)}
        \label{fig:global-vs-shared-3d}
    \end{minipage}
\end{figure}

@@

In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The Intel MKL solver can use multiple threads
to solve several systems simaltaneously.
The CUSPARSE solver uses a hybrid algorithm composed
of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).

The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.

Table \ref{table:bench} shows
timings for different solvers.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/bench-2d.eps}
\end{center}
\caption{Relative solver performance for 2-D problems}
\label{fig:bench-2d}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/bench-3d.eps}
\end{center}
\caption{Relative solver performance for 3-D problems}
\label{fig:bench-3d}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/phases-analysis.eps}
\end{center}
\label{fig:k20-forward-vs-backward}
\end{figure}

\begin{table}[]
\scriptsize
\centering
\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers}
\label{table:bench}
\input{table/bench.tex}
\end{table}

@@

\pagebreak
\pagebreak
\section*{References}

>>>>

\usepackage{amsmath,bm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{minted}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{rotating}

@@

\begin{frontmatter}
\author{Author names}
\title{A fast GPU solver for near-Toeplitz tridiagonal systems}
\maketitle

\begin{abstract}
    In this paper, we consider the solution of linear
    systems of the form $A\bm{x} = \bm{d}$,
    where $A$ is \emph{near-Toeplitz}.
    These systems appear in numerical schemes
    such as
    Alternating Direction Implicit (ADI) methods,
    compact finite differences,
    and one-dimensional ordinary
    and partial differential equations.
    We develop a GPU solver that takes advantage
    of the known structure of the matrix
    to reduce the number of computations
    and improve memory access.
    Our solver is able to achieve up to 5x speedup
    over an Intel MKL solver running on a single core,
    and 1.5x-11x speedup
    over the NVIDIA CUSPARSE solver.
    For larger problem sizes,
    our solver is able to outperform
    CUSPARSE as well as Intel MKL (on up to 16 cores).
    We also describe the implementation of a multi-GPU solver
    using MPI and CUDA.
    We describe two applications:
    2-D ADI applied to the heat equation on a single GPU,
    and computation of
    3-D compact finite differences on multiple GPUs.

@@

This paper is organized as follows:
Section \ref{sec:preliminaries}
describes algorithms for solving tridiagonal systems on GPUs.
The cyclic reduction algorithm in particular, is discussed in detail,
and issues with its implementation on GPUs is discussed.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:gpu-implementation}
discusses the implementation of the algorithm on GPUs.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=250pt]{img/cyclic-reduction-precomputing.eps}
\end{center}
\caption{Storing forward reduction coefficients
    from each reduction step $m$}
\label{fig:cyclic-reduction-precomputing}
\end{figure}

@@

In this section, we describe the implementation of a GPU solver
for solving a given near-Toeplitz tridiagonal system
for $N_{rhs}$ right hand sides.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.

Following the algorithm described in \ref{sec:proposed-algorithm},
the forward reduction coefficients
are first computed and loaded into global memory.
The right hand sides are stored in a single
contiguous region global memory.
In both approaches,
we map each of the $N_{rhs}$ tridiagonal systems
to a thread block,
and equations or \emph{indices} of a system
to individual threads within a block.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=400pt]{img/global-and-shared.eps}
\end{center}
\caption{Shared memory (top) and Global memory (bottom) implementations---
    multiple kernel launches allow us to dedicate appropriate number
    of threads per block in the global memory approach}
\label{fig:global-and-shared}
\end{figure}

@@

This implementation works entirely on the GPU's global memory.
Each right-hand side is mapped to a single thread block.
Each block executes two kernels,
\texttt{forwardReduction} and \texttt{backSubstitution},
which are called each called $log_2(n)-1$ times.
An extra call to the
forward reduction kernel additionally performs
the two-by-two solve.
Exit from a kernel marks a global synchronization point,
so no explicit synchronization is performed in the kernels.
Because we launch a kernel for each
forward reduction and back substitution step,
we can specify the appropriate number of threads per block---$n/2$
threads per block for the first forward reduction step,
$n/4$ for the second, and so on (Figure \ref{fig:global-and-shared}).
This guarantees that
all threads within a block are always active.
The precomputed coefficient arrays and right hand
are accessed by the kernels from global memory directly.
Because the coefficients
$k_{1,i}$ and $k_{2,i}$
required in the forward reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
are precomputed and available in global memory,
we do not compute
$a_i^\prime$, $b_i^\prime$, $c_i^\prime$,
$k_{1,i}$ and $k_{2,i}$.
This reduces the number of uncoalesced accesses to global memory.
In the backward substitution steps
(Equation \ref{eqn:precomputed-backward-substitution-step}),
the coefficients $a^m$, $b^m$ and $c^m$
are accessed from global memory without coalescing problems.

@@

In the shared memory approach (Figure \ref{fig:global-and-shared}),
we again map every right-hand side to a thread block.
Each thread block executes a single kernel \emph{once},
which performs forward reduction,
backward substitution, and the two-by-two solve.
Shared memory sized $n/2$ is allocated by each block.
Each thread performs the first reduction step
(Equation \ref{eqn:precomputed-forward-reduction-step})
by accessing the required values
$d_i$, $d_{i-1}$ and $d_{i+1}$ from global memory,
storing the result in shared memory.
In subsequent reduction steps,
$d_i$, $d_{i-1}$ and $d_{i+1}$
are accessed from shared memory,
avoiding uncoalesced global memory accesses.
In each back substitution step,
threads overwrite the existing values in shared memory
with the values of the solution.
In the final step,
shared memory is filled completeley
with the even-indexed solution values.
Each thread then computes an odd-indexed solution value,
storing it directly in global memory
and copies the even-indexed solution value
from shared memory to global memory.
The entire solution is done in a single kernel launch
to hide the latency of
memory transfers between global and shared memory.
Explicit synchronization between the threads of a block
is required within the kernel at the end of each step.

%----------------------------------------------------------------------%
\section{Results} \label{sec:results-single-gpu}
%----------------------------------------------------------------------%

In this section we present a performance overview
of our Near-Toeplitz tridiagonal solver (NEATO)
against a multi-threaded Intel MKL solver and
the CUSPARSE GPU solver
(\texttt{dgtsv} and \texttt{dgtsvStridedBatch} respectively).
The CPU code is compiled with the Intel C compiler
and run on up to 16 cores of the
Intel Xeon Processor E5-2670 v2 (25M Cache, 2.50 GHz).
GPU code is compiled with the CUDA toolkit (version 6.5.14),
and run on the
NVIDIA Tesla K20 Accelerator.
\texttt{-O2} level optimizations are turned on for both
CPU and GPU code.

@@

\begin{figure}[!htb]
        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
        \caption{Comparison of global memory and shared
            memory implementations of NEATO (2D problems)}
        \label{fig:global-vs-shared-2d}
        \includegraphics[width=1.0\linewidth]{fig/global-vs-shared-2d.eps}
        \caption{Comparison of global memory and shared memory
            implementations of NEATO (2D problems)}
        \label{fig:global-vs-shared-3d}
\end{figure}

@@

In Figures \ref{fig:bench-2d} and \ref{fig:bench-3d},
we provide the relative performance of
Intel MKL and CUSPARSE solvers and compare against
the NEATO shared memory implementation.
The Intel MKL solver can use multiple threads
to solve several systems simaltaneously.
The CUSPARSE solver uses a hybrid algorithm composed
of Cyclic Reduction and Parallel Cyclic Reduction (CR+PCR).
The relative performance for
each problem size is obtained by
normalizing the solver timings
by the timing for the fastest solver for that problem size.
Table \ref{fig:bench} shows the timings of the various
solvers to solve different problem sizes.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/bench-2d.eps}
\caption{Relative solver performance for 2-D problems}
\label{fig:bench-2d}
\end{center}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=300pt]{fig/bench-3d.eps}
\caption{Relative solver performance for 3-D problems}
\label{fig:bench-3d}
\end{center}
\end{figure}

\begin{sidewaystable}[]
\scriptsize
\centering
\caption{Performance of Intel MKL, CUSPARSE and NEATO solvers}
\label{table:bench}
\input{table/bench.tex}
\end{sidewaystable}

@@

\subsection{Numerical algorithm}

In a uniformly spaced one-dimensional grid with spacing $dx$,
if $f_i$ represents the value of
the function evaluated at the $i$th node,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:

\begin{equation}
\begin{split}
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
        f^{\prime}_i
    = 
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}

For the particular scheme used, $\alpha = 1/4$,
and $ a = 3/4 $.
It is easily noted that this leads to a tridiagonal system of equations,
along with a 3-point stencil to evaluate the RHS.
Further, the left boundary can be treated using the following implicit equation:

\begin{equation}
    f^{\prime}_1 + 2f^{\prime}_2 = \frac{-5f_1 + 4f_2 + f_3}{dx}
\end{equation}

The right boundary ($f^{\prime}_{n}$) is treated by utilizing
the negative complex-conjugate of the Fourier image of the stencil
at $f^{\prime}_1$:

\begin{equation}
    f^{\prime}_{n} + 2f^{\prime}_{n-1s}
    =
    \frac{5f_{n} - 4f_{n-2} 1 f_{n-2}}{dx}
\end{equation}

This boundary treatment maintains the tridiagonal identity
of the syste to be solved, which then has the form:

\begin{equation} \label{eqn:compact-tridiagonal-system}
 \begin{bmatrix}
     1&2\\
     1/4&1&1/4\\
     &1/4&1&1/4\\
     &&1/4&1&1/4\\
     &&&1/4&1&1/4\\
     &&&&&\ddots\\
     &&&&&&\ddots\\
     &&&&&&&\ddots\\
     &&&&&&&2&1
  \end{bmatrix}
  \begin{bmatrix}
      f^{\prime}_1 \\
      f^{\prime}_2 \\
      f^{\prime}_3 \\
      \vdots \\
      \vdots \\
      \vdots \\
      \vdots \\
      f^{\prime}_{n-1} \\
      f^{\prime}_n
   \end{bmatrix}
 =
 \begin{bmatrix}
     -5f_1 + 4f_2 + f_3\\
     f_{3} - f_{1}\\
     f_{4} - f_{2}\\
     \vdots\\
     \vdots\\
     \vdots\\
     \vdots\\
     f_{n} - f_{n-2}\\
     5f_{n} - 4f_{n-1} - f_{n-2}
  \end{bmatrix}
\end{equation}

\subsection{Computational domain}

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/computational-domain.eps}
\end{center}
\caption{Computational domain in 3-D}
\label{fig:computational-domain}
\end{figure}

The computational domain (Figure \ref{fig:computational-domain})
is a structured grid comprised of
$nx \times ny \times nz$ grid points.
While evaluating the derivative in, say, the x-direction,
the computational domain is thought of as
a collection of \emph{grid lines}---each consisting of
$nx$ grid points---oriented in the x-direction.
For each grid line,
we may assemble a tridiagonal system of size $nx$, as in
Equation \ref{eqn:compact-tridiagonal-system}.
Solving for the all derivatives in the $x$ direction then involves
solving this tridiagonal system for $ny \times nz$ right hand sides.

\subsection{Multi-GPU solver}

For multiple GPUs, we use the algorithm
proposed by Mattor et al. \cite{mattor1995algorithm}.
Here, we distribute the tridiagonal system among
$p$ procesors.
We solve for the ``local''  
portion of the global tridiagonal system
by first solving the following three ``local''
tridiagonal systems:

\begin{align}
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
x_{r,1}^p \\
x_{r,2}^p \\
x_{r,3}^p \\
x_{r,4}^p \\
\vdots \\
x_{r,m}^p
\end{bmatrix}
=
\begin{bmatrix}
r_1^p \\
r_2^p \\
r_3^p \\
r_4^p \\
\vdots \\
r_m^p
\end{bmatrix} & \label{eqn:primary-system} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
u_1^p \\
u_2^p \\
u_3^p \\
u_4^p \\
\vdots \\
u_m^p
\end{bmatrix}
=
\begin{bmatrix}
-a_1^p \\
0 \\
0 \\
0 \\
\vdots \\
0
\end{bmatrix} & \label{eqn:secondary-system-1} \\
%
%
%
& \begin{bmatrix}
b_1^p & c_1^p \\
a_2^p & b_2^p & c_2^p \\
      & a_3^p & b_3^p & c_3^p \\
      &       & a_4^p & b_4^p & c_4^p \\
      &       &       &       &  \ddots & c_{m-1}^p\\
      &       &       &       &     a_{m}^p  & b_{m}^p
\end{bmatrix}
\begin{bmatrix}
l_1^p \\
l_2^p \\
l_3^p \\
l_4^p \\
\vdots \\
l_m^p
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
0 \\
0 \\
\vdots \\
-c_m^p
\end{bmatrix} & \label{eqn:secondary-system-2}
\end{align}

and additionally, the following ``reduced'' system:

\begin{align} \label{eqn:reduced-system}
&
\begin{bmatrix}
l^1_m & -1 \\
-1    & u^2_1 & l^2_1 \\
      & u^2_m & l^2_m & -1 \\
      &       & -1    & u^3_1 & l^3_1 \\
      &       &       & u^3_m & l^3_n  & -1 \\
      &       &       &       & \ddots & \ddots & \ddots \\
      &       &       &       &        & -1     & u^P_1
\end{bmatrix}
\begin{bmatrix}
\beta^1 \\
\alpha^2 \\
\beta^2 \\
\alpha^3 \\
\beta^3 \\
\vdots \\
\alpha^P
\end{bmatrix}
=
\begin{bmatrix}
x_{r,m}^1 \\
x_{r,1}^2 \\
x_{r,m}^2 \\
x_{r,1}^3 \\
x_{r,m}^3 \\
\vdots \\
x_{r,1}^P \\
\end{bmatrix}
&
\end{align}

We call the system in Equation \ref{eqn:primary-system}
the ``primary'' system, and the systems in
Equations
\ref{eqn:secondary-system-1} and \ref{eqn:secondary-system-2}
the ``secondary'' systems.
The local part of the solution to the ``global'' tridiagonal system
is obtained as a linear combination of
the solutions to the primary and secondary sytems:

\begin{equation}
    \boldsymbol{x}^p = \boldsymbol{x}_r^p + \
        \alpha^p \boldsymbol{u}^p + \beta^p \boldsymbol{l}^p
    \label{eqn:sum-of-systems}
\end{equation}

Where $\alpha^p$ and $\beta^p$ are obtained
from the solution of the reduced system.

\subsection{GPU implementation}

\begin{enumerate}
    \item Computation of the RHS (requires communication)
    \item Solving the primary and secondary systems
    \item Constructing and solving the reduced systems,
        and scattering the paramters $\alpha^p$ and $\beta^p$ (requires communication)
    \item Summing the solutions (Equation \ref{eqn:sum-of-systems})
\end{enumerate}

Each of these is performed on the GPU,
one of the primary concerns is to eliminate
CPU-GPU transfers as much as possible in the solution process.

The right-hand side computation involves stencil computation,
which requires communication at the boundaries.
Communication of non-contiguous MPI
data types is expensive on the GPU,
so we have dedicated
contiguous halo arrays for performing swaps in each direction

\subsection{Results}

% Time for dfdx, dfdy, dfdz at various problem sizes
% and nprocs

% Breakdown - time allocated for each step

% Weak scaling

% Strong scaling

\pagebreak[4]
\section*{References}
